---
output: github_document
---

[![R-CMD-check](https://github.com/dongwenluo/safePivot/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dongwenluo/safePivot/actions/workflows/R-CMD-check.yaml)

[![Codecov test coverage](https://codecov.io/gh/dongwenluo/safePivot/branch/main/graph/badge.svg)](https://app.codecov.io/gh/dongwenluo/safePivot?branch=main)

# safePivot

`safePivot` is an R htmlwidget for interactive drag-and-drop pivot tables.

It wraps the original PivotTable.js library with safer defaults for data analysis:

- robust numeric aggregators
- missing-value handling
- R factor-level ordering
- heatmap renderers
- readable conditional formatting
- Shiny config capture
- CSV, Excel, RDS, and RData export helpers

## Installation

```{r, eval=FALSE}
# Development version
install.packages("remotes")
remotes::install_github("dongwenluo/safePivot")
```

## Basic use

```{r, eval=FALSE}
library(safePivot)

safePivot(
  iris,
  rows = "Species",
  vals = "Sepal.Length",
  aggregator = "Median",
  renderer = "Table"
)
```

## Heatmap

```{r, eval=FALSE}
safePivot(
  mtcars,
  rows = "cyl",
  cols = "gear",
  vals = "mpg",
  aggregator = "Mean",
  renderer = "Heatmap",
  heatmap_palette = "blue",
  conditional_format = TRUE
)
```

## Factor order

```{r, eval=FALSE}
iris2 <- iris
iris2$Species <- factor(
  iris2$Species,
  levels = c("virginica", "versicolor", "setosa")
)

safePivot(
  iris2,
  rows = "Species",
  vals = "Sepal.Length",
  aggregator = "Median",
  renderer = "Table",
  respect_factor_order = TRUE
)
```

## Missing values

```{r, eval=FALSE}
iris_missing <- iris
iris_missing$Group <- iris_missing$Species
iris_missing$Group[c(1, 5, 10)] <- NA
iris_missing$Sepal.Length[c(2, 6, 11)] <- NA

safePivot(
  iris_missing,
  rows = "Group",
  vals = "Sepal.Length",
  aggregator = "N missing",
  renderer = "Table",
  missing_label = "(Missing)"
)
```

## Shiny export pattern

```{r, eval=FALSE}
library(shiny)
library(safePivot)

ui <- fluidPage(
  safePivotOutput("pivot", height = "600px"),
  downloadButton("download_csv", "Download CSV"),
  downloadButton("download_xlsx", "Download Excel")
)

server <- function(input, output, session) {
  dat <- reactive(iris)

  output$pivot <- renderSafePivot({
    safePivot(
      dat(),
      rows = "Species",
      vals = "Sepal.Length",
      aggregator = "Median",
      renderer = "Table"
    )
  })

  current_pivot <- reactive({
    safe_pivot_compute_from_config(
      data = dat(),
      config = input$pivot_config,
      default_rows = "Species",
      default_vals = "Sepal.Length",
      default_aggregator = "Median"
    )
  })

  output$download_csv <- downloadHandler(
    filename = function() "safePivot_result.csv",
    content = function(file) {
      safe_pivot_write_csv(current_pivot(), file, which = "wide")
    }
  )

  output$download_xlsx <- downloadHandler(
    filename = function() "safePivot_result.xlsx",
    content = function(file) {
      safe_pivot_write_xlsx(current_pivot(), file)
    }
  )
}

shinyApp(ui, server)
```

## Aggregators

Common aggregators include:

```{r, echo=FALSE}
data.frame(
  Type = c(
    "Count",
    "Missingness",
    "Numeric summary",
    "Distribution",
    "Fraction"
  ),
  Examples = c(
    "Count, Count unique, List unique values",
    "N missing, N non-missing, Missing %, Non-missing %",
    "Sum, Mean, Median, Min, Max, Range",
    "Q1, Q3, IQR, Variance, SD, SE, CV %",
    "Sum as Fraction of Total, Count as Fraction of Rows"
  )
)
```

## Notes

`safePivot` is designed for interactive exploration. For very large data, pre-filter or aggregate data before sending it to the browser.

## Related work and acknowledgements

`safePivot` is inspired by `rpivotTable`, an R htmlwidget that brings
PivotTable.js / pivottable.js drag-and-drop pivot tables to R.

`safePivot` takes a more focused and conservative direction: it keeps the
drag-and-drop table workflow, but removes chart renderers and adds extra
safeguards for numeric aggregation, missing values, factor-level ordering,
readable heatmaps, Shiny config capture, and R-side CSV/Excel/RDS/RData export.

We also acknowledge PivotTable.js by Nicolas Kruchten, the underlying
JavaScript pivot-table library used for the interactive browser-side pivot UI.

