---
title: "Getting started with safePivot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with safePivot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

`safePivot` provides a safe drag-and-drop pivot table htmlwidget for R and Shiny. It wraps PivotTable.js with robust numeric aggregators, missing-value summaries, zero-value summaries, factor-order preservation, heatmaps, lightweight Plotly chart renderers, Shiny configuration capture, and export helpers.

The package is designed for data exploration workflows where users need to distinguish:

- empty row-column combinations
- observed missing values
- observed zero values
- positive and non-zero measured values

This distinction is important for scientific and agricultural datasets, where a missing measurement, a zero measurement, and a non-existing row-column combination have different meanings.

# Basic pivot table

```{r basic-compute}
library(safePivot)

res <- safe_pivot_compute(
  iris,
  rows = "Species",
  vals = "Sepal.Length",
  aggregator = "Median"
)

res$wide
```

# Interactive widget

The interactive widget is usually run in the RStudio Viewer, a browser, R Markdown, or Shiny.

```{r basic-widget, eval=FALSE}
safePivot(
  iris,
  rows = "Species",
  cols = "Petal.Width",
  vals = "Sepal.Length",
  aggregator = "Median",
  renderer = "Table"
)
```

# Heatmap

SafePivot supports table-style heatmaps:

```{r heatmap-widget, eval=FALSE}
safePivot(
  iris,
  rows = "Species",
  cols = "Petal.Width",
  vals = "Sepal.Length",
  aggregator = "Median",
  renderer = "Heatmap",
  heatmap_palette = "green"
)
```

Heatmap renderers include:

```r
"Heatmap"
"Row Heatmap"
"Col Heatmap"
```

Browser-side conditional formatting is intentionally disabled for heatmap renderers so that the heatmap palette is not overwritten by SafePivot cell-colour classes.

# Lightweight Plotly charts

SafePivot includes lightweight Plotly chart renderers:

```{r plotly-widget, eval=FALSE}
safePivot(
  iris,
  rows = "Species",
  cols = "Petal.Width",
  vals = "Sepal.Length",
  aggregator = "Mean",
  renderer = "Bar Chart",
  plot_height = 620
)
```

Supported lightweight chart renderers include:

```r
"Horizontal Bar Chart"
"Horizontal Stacked Bar Chart"
"Bar Chart"
"Stacked Bar Chart"
"Line Chart"
"Area Chart"
"Scatter Chart"
"Multiple Pie Chart"
```

Treemap is intentionally not included in the core package because it requires a larger Plotly bundle and custom hierarchy handling.

# Missing values and zero values

SafePivot includes data-quality aggregators:

```r
"N missing"
"N non-missing"
"Missing % within Cell"
"Non-missing % within Cell"
"N zero"
"N non-zero"
"Zero % within Cell"
"Non-zero % within Cell"
```

Example data:

```{r missing-data}
dat <- data.frame(
  treatment = c("A", "A", "B", "B", "C", "C"),
  season = c("Spring", "Summer", "Spring", "Summer", "Spring", "Summer"),
  yield = c(10, NA, 0, 12, 15, NA)
)

safe_pivot_compute(
  dat,
  rows = "treatment",
  cols = "season",
  vals = "yield",
  aggregator = "N missing"
)$wide
```

Zero values are counted separately from missing values:

```{r zero-data}
safe_pivot_compute(
  dat,
  rows = "treatment",
  cols = "season",
  vals = "yield",
  aggregator = "N zero"
)$wide
```

# Fraction aggregators

SafePivot includes sum and count fraction aggregators:

```r
"Sum as Fraction of Total"
"Sum as Fraction of Rows"
"Sum as Fraction of Columns"
"Count as Fraction of Total"
"Count as Fraction of Rows"
"Count as Fraction of Columns"
```

For example:

```{r count-fraction}
safe_pivot_compute(
  iris,
  rows = "Species",
  vals = "Sepal.Length",
  aggregator = "Count as Fraction of Total"
)$wide
```

Each species is one third of the `iris` dataset.

# Factor ordering

SafePivot can respect R factor-level ordering.

```{r factor-order}
dat2 <- data.frame(
  treatment = factor(
    rep(c("Control", "Low", "Medium", "High"), each = 4),
    levels = c("Control", "Low", "Medium", "High")
  ),
  season = factor(
    rep(c("Spring", "Summer", "Autumn", "Winter"), times = 4),
    levels = c("Spring", "Summer", "Autumn", "Winter")
  ),
  yield = c(10, 11, 12, 13, 14, 15, 16, 17,
            18, 19, 20, 21, 22, 23, 24, 25)
)

safe_pivot_compute(
  dat2,
  rows = "treatment",
  cols = "season",
  vals = "yield",
  aggregator = "Median"
)$wide
```

For the interactive widget:

```{r factor-order-widget, eval=FALSE}
safePivot(
  dat2,
  rows = "treatment",
  cols = "season",
  vals = "yield",
  aggregator = "Median",
  renderer = "Table",
  respect_factor_order = TRUE
)
```

# Export

SafePivot can export computed pivot results.

```{r export-examples, eval=FALSE}
res <- safe_pivot_compute(
  iris,
  rows = "Species",
  cols = "Petal.Width",
  vals = "Sepal.Length",
  aggregator = "Median"
)

safe_pivot_write_csv(res, "iris_pivot.csv")
safe_pivot_write_rds(res, "iris_pivot.rds")
safe_pivot_write_rdata(res, "iris_pivot.RData")

# Requires openxlsx
safe_pivot_write_xlsx(res, "iris_pivot.xlsx")
```

# Development checks

Before releasing updates, run:

```r
devtools::document()
devtools::load_all(reset = TRUE)
devtools::test()
devtools::check()
```